---
title: "Rapid EDA"
format: 
    html: 
       code-fold: true
       code-tools: true
execute: 
  warning: false
  message: false
---

A first step is to rapidly evaluate raw data.

In creating regional health indicators and profiles

Store data in a single directory

```{r}
#| label: setup
#| include: false

library(needs)
needs(tidyverse, data.table, here, fs, readxl)

```

```{r}
#| label: load datasets
#| output: false

dir <- here("data")

xl_files <- fs::dir_ls(dir, regexp = "xls")

csv_files <- fs::dir_ls(dir, regexp = "csv")

## read_files 

xl <- map(xl_files, read_xlsx)
csvs <- map(csv_files, read_csv)

```

```{r}
#| label: region names
#| output: false

map(xl, colnames)
map(csvs, colnames)

```

| Dataset     | Area field name    |
|-------------|--------------------|
| AMR         | No area variable   |
| Injury      | `Region`           |
| Flu         | `region_en`        |
| Smoking     | `directorate_name` |
| Populations | `Region`           |

: Area name labels

To facilitate data linkage and creating indicator datasets, area variable names should be consistent between datasets.

Directorate is not equivalent to region.

There are 13 KSA regions and 20 health directorates

## Area variable names

```{r}
#| label: area variable names
#| output: false

## rename area variables

csvs$`/Users/julianflowers/poc/data/Flu Vaccine Coverage 2023 updated.csv` <- rename(csvs$`/Users/julianflowers/poc/data/Flu Vaccine Coverage 2023 updated.csv`, Region = region_en)

#csvs$`/Users/julianflowers/poc/data/Flu Vaccine Coverage 2023 updated.csv`

```

## Area names

```{r}
#| label: area names

flu_areas <- csvs$`/Users/julianflowers/poc/data/Flu Vaccine Coverage 2023 updated.csv` |>
    select(Region) |> unique()

smoking_areas <- csvs$`/Users/julianflowers/poc/data/Smoking 2022.csv` |>
    select(directorate_name) |> unique()

pop_areas <- csvs$`/Users/julianflowers/poc/data/Translated_Population_Data_with_Governorates.csv` |>
    select(Region) |> unique()

injury_areas <- xl$`/Users/julianflowers/poc/data/Nonfatal Hospitalizations for Injuries data 2023 (8-7-2024).xlsx` |> select(Region) |> unique()

n_areas <- data.frame(data = c("flu_areas", "smoking_areas", "pop_areas", "injury_areas"), no_areas = map_dbl(list(flu_areas, smoking_areas, pop_areas, injury_areas), nrow), area_type = c("region", "directorate", "region", "region")) |>
    knitr::kable()
    

```

The number of unique areas `r n_areas`

```{r}
#| label: area name overap

setdiff(flu_areas, pop_areas)
setdiff(pop_areas, injury_areas)
setdiff(injury_areas, flu_areas)

```

