---
title: "Health profile (2)"
subtitle: "Maps"
format: 
    html: 
        fig-width: 8
        fig-height: 8
        fig-cap: bottom
execute: 
  freeze: true
---

```{r}
#| label: setup
#| include: false

needs(tidyverse, mapview, sf)

## load boundary file

ksa_bound <- read_sf("data/ksa_bound.gml")


```

### Boundary file

A regional boundary file is downloaded as part of the data pre-processing step and is contained in the object `sa_bound`.

This can be plotted as a static or interactive map:

```{r}
#| label: fig-basic-map
#| fig-cap: "KSA boundary map"
#| fig-subcap: 
#|     - "static"
#|     - "interactive"

ksa_bound |>
    ggplot() +
    geom_sf()

mapview(ksa_bound)

```

### Plot data

Attribute data (e..g area rates) needs to be combined with the boundary data in order to map them. Taking the example of injury....

```{r}

inj <- read_csv("data/injury_dsr.csv")


```

### Rename regions to link to spatial file

```{r}

setdiff(inj$region, ksa_bound$ADM1_EN)
  
ksa_bound$ADM1_EN 

## recode area names and join to boundary data
inj_sf <- inj |>
    mutate(region1 = recode(region, 
                           "'Asir" = "`Asir", 
                           "Al Madinah al Munawwarah" = "Al Madinah",
                           "Al Qasim" = "Al Quassim", 
                           "Makkah al Mukarramah" = "Makkah", 
                           "Al Mintaqah ash Sharqiyah" = "Ash Sharqiyah", 
                           "Ar Riyadh" = "Ar Riyad", 
                           "Jazan" = "Jizan"), 
           region1 = ifelse(region1 != region, region1, region)
           )

inj_sf <- left_join(ksa_bound, inj_sf,  by = c("ADM1_EN" = "region1")) 

inj_sf |>
    filter(is.na(gender))
    


```

### Plot smoking rates as choropleth map

```{r}
#| label: fig-sm-map
#| fig-cap: "Smoking rate map"
#| fig-subcap: 
#|      - "Static"
#|      - "Interactive"

## convert back to spatial data and plot
inj_sf |> st_as_sf() |>
    filter(!is.na(gender)) |>
    ggplot() +
    geom_sf(data = ksa_bound) +
    geom_sf(aes(fill = 100000 * adj.rate)) +
    ggplot2::geom_sf_label(aes(label = ADM1_EN)) +
    facet_wrap(~ gender, ncol = 1) +
    theme_void() +
    scale_fill_viridis_c(name = "DSR Injury\nper 100,000")

mapview(inj_sf |> st_as_sf() |>
    filter(gender == "female") |> mutate(adj.rate = 100000 * adj.rate), zcol = "adj.rate") 



```
