---
title: "Indicator definitions"
format: 
    html:
        fig-height: 10
        fig-width: 8
        toc: true
        toc-location: right
        code-fold: true
        code-tools: true
execute:
    freeze: false
    echo: false
---

```{r}
#| label: setup 
#| include: false

needs(tidyverse, flextable, gt)

indicators <- read_csv("data/indicators.csv")



```

These are as extracted from the definitions as agreed for the proof of concept

## Anti-microbial resistance

Percentage of bloodstream infections due to selected antimicrobial-resistant organisms (MRSA, Escherichia coli resistant to 3rd-generation cephalosporin (e.g., ESBL- E. coli))

### Staphylococcus aureus

```{r}
#| label: staph areus

amr_def_sa <- data.frame(
    title = "Percentage of bloodstream infections due to selected antimicrobial-resistant organisms: MRSA", 
    denominator = "Total No. of patients with growth of S. aureus in tested blood samples", 
    numerator = "No. of patients with growth of methicillin-resistant S. aureus in tested blood samples", 
    metric = "Proportion", 
    notes = ""
) 

amr_def_sa |>
    flextable(cwidth = 6)

```

### Escherichia coli

```{r}
#| label: e coli

amr_def_ec <- data.frame(
    title = "Percentage of bloodstream infections due to selected antimicrobial-resistant organisms: E.coli", 
    denominator = "Total No. of patients with growth of E. coli in tested blood samples", 
    numerator = "No. of patients with growth of E.coli resistant to 3rd generation cephalosporins in tested blood samples", 
    metric = "Proportion", 
    notes = "Resistance to ceftazidine, cefotaxine"
) 

amr_def_ec |>
    flextable(cwidth = 6)

```

## Smoking

Current cigarette smoking among women aged 18-44 years

```{r}
#| label: smoking

smok_def <- data.frame(
    title = "PCurrent cigarette smoking among women aged 18-44 years", 
    denominator = "Women aged 18-44 years who information about cigarette smoking (excluding unknowns and refusals).", 
    numerator = "Women aged 18-44 years who reported that they smoked ≥100 cigarettes in their lifetime and currently smoke every day or some days.", 
    metric = "Rate with 95% confidence intervals", 
    notes = "Data provided is attendances at smoking clinics rather than survey based smoking prevalence"
) 

smok_def |>
    flextable(cwidth = 6)

```

## Injury

Nonfatal hospitalizations for all injuries

```{r}
#| label: injury

inj_def <- data.frame(
    title = "Nonfatal hospitalizations for all injuries", 
    denominator = "Midyear population for the calendar year under surveillance obtained from the KSA Census or suitable alternative.", 
    numerator = "Nonfatal hospitalizations with any of the following ICD-10-CM diagnostic codes in the principal diagnosis field", 
    metric = "Rate (age standardised) with 95% confidence intervals", 
    notes = "Data provided is attendances at smoking clinics rather than survey based smoking prevalence"
) 

inj_def |>
    flextable(cwidth = 6)

```

## Flu vaccination coverage

### Paediatric

```{r}
#| label: flu paediatric

flu_def <- data.frame(
    title = "Population vaccination coverage - Flu (pediatric group 6 months - 18 years old))", 
    denominator = "The total number of children eligible for influenza vaccination",
    numerator = "The total number of children in the respective eligible age cohort that have received at least one dose of influenza vaccine from 1 Oct in school, pharmacy, and general practice.", 
    metric = "Proportion with 95% confidence intervals", 
    notes = "Data provided is attendances at smoking clinics rather than survey based smoking prevalence"
) 

flu_def |>
    flextable(cwidth = 6)


```

### At-risk groups

The dataset required for calculating coverage for at-risk groups is much more complex than for age-specific coverage rates because it needs details of underlying diseases e.g. asthma, COPD.

### 65+

```{r}

#| label: flu 65 and over

flu_def_65 <- data.frame( title = "Population vaccination coverage - Flu (65+))", denominator = "The total number of adults 65+ eligible for influenza vaccination", numerator = "The total number of children in the respective eligible age cohort", metric = "Proportion with 95% confidence intervals", notes = "Data provided is attendances at smoking clinics rather than survey based smoking prevalence" )

flu_def_65 |> flextable(cwidth = 6)

```
