---
title: "Health profile (1)"
execute: 
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false

needs(tidyverse)
options(scipen = 999)

## load smoking data and clean fields
smok_rates <- read_csv("data/smok_rates.csv") |>
    mutate(age = "18-44", 
           indicator = paste("smoking", age)) |>
    select(indicator, region = Region, age, gender = Gender.x, n = smokers, pop = sumpops, value, lowercl, uppercl) |> mutate(gender = recode(gender, "Female" = "female")) |>
    mutate(timeperiod = 2023)

flu_coverage <- read_csv("data/flu_coverage.csv") |>
    mutate(indicator = paste("flu_coverage", age)) |>
    rename(region = Region, gender = Gender)

smok_rates <- smok_rates |>
    add_row(
        data.frame(
            indicator = "smoking", 
            region = "KSA", 
            age = "18-44",
            gender = "female", 
            n = sum(smok_rates$n, na.rm = TRUE), 
            pop = sum(smok_rates$pop),
            value = 100000 * sum(smok_rates$n, na.rm = TRUE) / sum(smok_rates$pop),
            lowercl = NA, 
            uppercl = NA
        )
    ) |>
    mutate(timeperiod = 2023) 

```

## Putting it all together

Health profiles combine multiple indicators to allow comparison, usually between population or administrative units e.g. regions, hospitals, ethnic groups, age groups and provide tools to facilitate a range of comparison.

### Workflow

#### Preparing data

To generate profile-style presentations per-indicator datasets need to be configured in a standard way and pooled. The minimum meta-data (variable names) for each indicator would be:

| Indicator name                    |
|-----------------------------------|
| Area name                         |
| Age group                         |
| Gender                            |
| Value                             |
| Count (numerator)                 |
| Denominator                       |
| Metric type e.g. proportion, rate |
| Unit e.g. per 100,000, %          |
| Lower uncertainty interval        |
| Upper uncertainty interval        |
| Uncertainty e.g. 95%, 99.8%       |

: Meta-data for profile data

```{mermaid}
%%| label: fig-workflow-prof
%%| fig-cap: "Proposed workflow for creating health profiles"

flowchart TD
A[Configure per-indicator data] --> B[bind datasets]



```

UK health profiles (for example, Public Health Outcome Framework) offer a wide range of comparison:

-   Between and within areas

-   Between age groups and genders

## Visualisation

| Comparison                            | Visualisation                         | Comment                                                                                                                                                                                                     |
|---------------|---------------|------------------------------------------|
| Single area, multiple indicators      | Spine or bullet chart                 | Indicator values are scaled so that they can be directly compared. Single area (unit) values are usually encoded as a dot or line, plotted on a bar representing the range of values for all areas / units. |
| Multiple areas, single indicator      | Column / bar chart                    | Plots can be ordered and plotted horizontally to enable rapid appraisal of relative performance. Confidence intervals can be added                                                                          |
| Indicator pairs / multiple indicators | Scatter plot or scatter plot matrices | May include smooths (lines representing fit to the data - common options are linear, loess and generalised additive models)                                                                                 |
| Spatial variation                     | Choropleth maps                       | Requires boundary files and GIS tools                                                                                                                                                                       |
| Overview                              | Heatmap / tile chart                  | Colour cells with statistical significance categories (e.g. high / low)                                                                                                                                     |
|                                       |                                       |                                                                                                                                                                                                             |

We will take the toolkit for plotting provided by the UKHSA which generates the charts used in the online version if the PHOF as a basis for plotting profiles. This is available at https://github.com/ukhsa-collaboration/fingertipscharts.

Note: in the PHOF statistical significance is defined by whether the global average value (e.g. KSA) is contained with the confidence limits of the unit values (e.g. 'Asir region). It does not calculate limits for the global value. The rationale for this is explained in @Public-Health-Data-Science2018-et

```{r}
## install fingertipscharts
## 

if(!require("fingertipscharts")) devtools::install_github("https://github.com/ukhsa-collaboration/fingertipscharts", force = TRUE)

library(fingertipscharts)

```

To use `fingertipscharts` data has to be in the right format.

Bar charts can be created using the `compare_areas` function

```{r}
#| label: fig-compare
#| fig-cap: "Using `compare-areas"
#| fig-subcap: 
#|     - "Smoking rates"
#|     - "Paediatric flu coverage"

smok_rates |>
    compare_areas(area = region, value = value, lowerci = lowercl, upperci = uppercl, title = "Rate of attendance at smoking clinics 18-44: F")

flu_coverage |>
    filter(gender == "female" & age == "0-18") |>
    compare_areas(area = region, value = value, lowerci = lowercl, upperci = uppercl, fill = gender, title = "Flu vaccination coverage 0-18; F")

```

## Data overview

PHOF uses a heatmap (known as a "tartan rug" to display area by indicator summaries. The term tartan rug is used to refer to the colour scheme (red-amber-green) to denote statistical significance. The

```{r}
#| label: tbl-format
#| tbl-cap: "Required data format example"


prof_data <- bind_rows(smok_rates, flu_coverage) |>
    select(-c(11:13)) 

## calculate ksa values
ksa_values <- prof_data |>
    group_by(indicator) |>
    reframe(ksa_values = 100000 * sum(n, na.rm = TRUE) / sum(pop))

## join to data
prof_data <- prof_data |>
    left_join(ksa_values)

## code for statistical significance

prof_data <- prof_data |>
    mutate(significance = case_when(lowercl > ksa_values ~"high" ,
                                     uppercl < ksa_values~ "low",
                                    TRUE ~ "ns"))

head(prof_data) |>
    gt::gt()

```

### Heatmap

```{r}

prof_data |>
    ggplot() +
    geom_tile(aes(region, indicator, fill = significance)) +
    scale_fill_manual(values = c("red", "darkgreen", "goldenrod")) +
    scale_x_discrete(position = "top") +
    theme(axis.text.x = element_text(angle = 90, hjust = 0)) +
    labs(y = "", x = "")
    

```

### 
