---
title: "Smoking"
format: html
editor: visual
execute: 
  warning: false
---

```{r}
#| label: setup
#| include: false

needs(tidyverse, data.table, googledrive, readxl, skimr, janitor, gt, gtsummary, collapsibleTree, PHEindicatormethods, ggthemes, sf, tmap, ggspatial, curl, raster, stars)

theme_set(ggthemes::theme_economist())

smoking <- read_csv("~/proof-of-concept/data/smoking.csv") |> setDT()
pops <- read_csv("~/proof-of-concept/data/pops.csv")

```

## Workflow

```{mermaid}
%%| label: fig-workflow
%%| fig-cap: "Proposed workflow for calculating smoking rates"

flowchart TD
A[Data] --> B[population data]
A --> C[smoking data]
C --> D[[reclassify directorate names to regions]] --> E[[calculate age bands]] --> F[[calculate smoking counts]]
B --> G[[calculate age bands]] --> H[[calculate population counts]]
H --> I[[join datasets]]
F --> I
I --> J[calculate rates and ci]

```

### Load census population and create age bands

```{r}
#| label: smoking pop data

## create age band

pops <- setDT(pops)[, `:=` (`18-44`  = dplyr::between(age, 18, 44), `15+` = age >= 15, `80+` = age >= 80, age_band = cut(age, seq(0, 110, 5), right = FALSE))][]

pop_age <- pops[, .(n = .N, sumpops = sum(Population, na.rm = TRUE)), by = .(Region, age_band,  Gender)][order(age_band)]

pop1844F <- pops[, .(n = .N, sumpops = sum(Population, na.rm = TRUE)), by = .(Region, `18-44`,  Gender)][`18-44` == "TRUE" & Gender == "Female",]

pop15F <- pops[, .(n = .N, sumpops = sum(Population, na.rm = TRUE)), by = .(Region, `15+`,  Gender)][`15+` == "TRUE" & Gender == "Female",]

```

### Load smoking data and calculate age bands

Note: it is not clear if the dummy data is a random sample of clinic attendance data.

```{r}
#| label: create age bands - 5 yr, 18-44, 15+


smok_age <- smoking[, `:=` (`18-44`  = dplyr::between(age, 18, 44), `15+` = age >= 15, `80+` = age >= 80, age_band = cut(age, seq(0, 110, 5), right = FALSE))][]

smok_age_bands <- smok_age[, .(n = .N, smokers = sum(n, na.rm = TRUE)), by = .(Region, age_band,  Gender)][order(age_band)]

smok1844F <- smok_age[, .(n = .N, smokers = sum(n, na.rm = TRUE)), by = .(Region, `18-44`,  Gender)][`18-44` == "TRUE" & Gender == "female",]

smok15F <- smok_age[, .(n = .N, smokers = sum(n, na.rm = TRUE)), by = .(Region, `15+`,  Gender)][`15+` == "TRUE" & Gender == "female",]

```

### Join datasets and calculate rates

This step uses Byar's method for confidence interval for rates

```{r}
#| label: fig-rate
#| fig-cap: "Smoking clinic attendance rates (18-44)"

smok_rates <- pop1844F |>
    left_join(smok1844F, by = "Region") |> 
    phe_rate(x = smokers, n = sumpops) 

smok_rates |>
    dplyr::select(Region, smokers, population = sumpops, value, lowercl, uppercl) |>
    flextable::flextable()

smok_rates |>
    write_csv("data/smok_rates.csv")
   
smok_rates |>
    ggplot() +
    geom_col(aes(reorder(Region, value), value)) +
    geom_linerange(aes(x = Region, ymin = lowercl, ymax = uppercl)) +
    coord_flip() +
    labs(x = " ",
         y = "Rate per 100.000")
    

```

### Age-specific attendance rates

```{r}
smok_age_agg <- smok_age_bands[, .(tot_smokers = sum(smokers)), by = .(age_band, Gender)][]

options(scipen = 999, digitds = 2)

## m:f as smoking ratios

smok_rates_gender <- pop_age |>
    dplyr::select(-Region) |>
    mutate(Gender = recode(Gender, "Female" = "female", "Male" = "male")) |>
    left_join(smok_age_agg, by = c("age_band", "Gender")) |>
    group_by(age_band, Gender) |>
    reframe(nat_smok = sum(tot_smokers, na.rm = TRUE), 
           nat_pop = sum(sumpops)) |>
    phe_rate(x = nat_smok, n = nat_pop) |> 
    mutate(value = ifelse(Gender == "female", -value, value)) |>
    ggplot() +
    geom_col(aes(age_band, value, fill = Gender)) +
    coord_flip() +
    scale_fill_manual(values = c("blue", "red")) +
    labs(x = "",  y = "Attendance rate per 100000")

smok_rates_gender    

smok_gender_ratio <- pop_age |>
    dplyr::select(-Region) |>
    mutate(Gender = recode(Gender, "Female" = "female", "Male" = "male")) |>
    left_join(smok_age_agg, by = c("age_band", "Gender")) |>
    group_by(age_band, Gender) |>
    reframe(nat_smok = sum(tot_smokers, na.rm = TRUE), 
           nat_pop = sum(sumpops)) |>
    phe_rate(x = nat_smok, n = nat_pop) |>
    slice(7:36) |>
    group_by(age_band) |>
    reframe(ratio = value[2] / value[1]) |>
    mutate(mean_ratio = mean(ratio))

```

The relative attendance rate for males is `r round(smok_gender_ratio$mean_ratio[1],2)` times that of females. Smoking clinic attendance rates could be used as a proxy for population smoking rates if the relationship between attendance and prevalenceis known . For example, survey data suggests that KSA male smoking rates are around 20% which implies a female smoking rate based of `r round(20/smok_gender_ratio$mean_ratio[1], 2)`%. In fact, available survey data, and estimates from the Global BUrden of Disease, suggest a smoking prevalence in females over 15 of 2%,
